---
title:  "Admin resources"
---

```{r,child="assets/header-lab.Rmd"}
```

```{r,eval=TRUE,include=FALSE}
library(yaml)
upid <- yaml::read_yaml("_site.yml")$uppmax_project
upres1 <- yaml::read_yaml("_site.yml")$uppmax_res_1
upres2 <- yaml::read_yaml("_site.yml")$uppmax_res_2
upres3 <- yaml::read_yaml("_site.yml")$uppmax_res_3
upres4 <- yaml::read_yaml("_site.yml")$uppmax_res_4
upres5 <- yaml::read_yaml("_site.yml")$uppmax_res_5
```

<div class="boxy boxy-lightbulb">
Useful utility commands/scripts for course managers.
</div>

# UPPMAX

Useful commands for monitoring user activity on UPPMAX.

## List jobs

List all jobs running an a project

```{r,echo=FALSE,comment="",class.output="bash"}
cat(paste0("squeue -A ",upid," | sort -k 4"))
```

Sample output

```
  23186770      core _interac analopez  R    2:15:00      1 r55
  23175660      core   (null)   anitav  R    6:52:28      1 r480
  23188713      core   (null)    annha  R    1:42:59      1 r480
  23185442      core   (null)    annha  R    2:47:11      1 r480
  23185433      core   (null) inene424  R    2:47:42      1 r479
  23185391      core   (null) inene424  R    2:50:04      1 r479
  23185540      core   (null)  shengyz  R    2:46:20      1 r480
     JOBID PARTITION     NAME     USER ST       TIME  NODES NODELIST(REASON)
  23185119      core   (null)    wingf  R    2:53:40      1 r479
```

## Core usage

Total number of cores used on a project

```{r,echo=FALSE,comment="",class.output="bash"}
cat(paste0("squeue -A ",upid," -o %C | awk '{total += $0} END{print total}'"))
```
## Reservation use

List users by reservation ID

```{r,echo=FALSE,comment="",class.output="bash"}
cat(paste0("squeue -R ",upres1," | sort -k 4\n"))
cat(paste0("squeue -R ",upres2," | sort -k 4\n"))
cat(paste0("squeue -R ",upres3," | sort -k 4\n"))
cat(paste0("squeue -R ",upres4," | sort -k 4\n"))
cat(paste0("squeue -R ",upres5," | sort -k 4\n"))
```
Sample output

```
23188713      core   (null)    annha  R    2:01:15      1 r480
23185442      core   (null)    annha  R    3:05:27      1 r480
23187840      core   (null)   evangc  R    2:11:02      1 r479
23186811      core   (null)  shengyz  R    2:34:45      1 r479
23185540      core   (null)  shengyz  R    3:04:36      1 r480
   JOBID PARTITION     NAME     USER ST       TIME  NODES NODELIST(REASON)
23185119      core   (null)    wingf  R    3:11:56      1 r479

```

## User activity

List last activity in a directory for all users

```{r,echo=FALSE,comment="",class.output="bash"}
cat(paste0("bash /sw/courses/utils/list_modification_times.sh /proj/",upid,"/nobackup/"))
```
Sample output

```
hkyle           2021-11-22 13:59:59     (/proj/snic2021-22-644/nobackup/hkyle/uppmax_tutorial/job_template)
malinh          2021-11-22 14:12:10     (/proj/snic2021-22-644/nobackup/malinh/slurm-23182638.out)
aliraz          2021-11-22 14:23:16     (/proj/snic2021-22-644/nobackup/aliraz/slurm-23184022.out)
kristaps        2021-11-22 14:25:08     (/proj/snic2021-22-644/nobackup/kristaps/uppmax_tutorial/jobData.sam)
anapin          2021-11-22 14:56:14     (/proj/snic2021-22-644/nobackup/anapin/uppmax_tutorial/uppmax_tutorial/jobData.sam)
analopez        2021-11-22 14:56:16     (/proj/snic2021-22-644/nobackup/analopez/uppmax_tutorial/jobData.sam)
private         Not available     ()
```

## List reservations for a project

```{r,echo=FALSE,comment="",class.output="bash"}
cat(paste0("scontrol show res | grep ",upid))
```

Sample output

```
ReservationName=snic2021-22-644_wed StartTime=2021-11-24T12:00:00 EndTime=2021-11-24T18:00:00 Duration=06:00:00
   Users=(null) Accounts=snic2021-22-644 Licenses=(null) State=INACTIVE BurstBuffer=(null) Watts=n/a
ReservationName=snic2021-22-644_thu StartTime=2021-11-25T08:30:00 EndTime=2021-11-25T17:30:00 Duration=09:00:00
   Users=(null) Accounts=snic2021-22-644 Licenses=(null) State=INACTIVE BurstBuffer=(null) Watts=n/a
ReservationName=snic2021-22-644_fri StartTime=2021-11-26T08:30:00 EndTime=2021-11-26T13:30:00 Duration=05:00:00
   Users=(null) Accounts=snic2021-22-644 Licenses=(null) State=INACTIVE BurstBuffer=(null) Watts=n/a
```

***
