---
title: "Schedule"
output:
  bookdown::html_document2:
    toc: false
    number_sections: false
---

```{r,child="assets/header-lab.Rmd"}
```

```{r,eval=TRUE,include=FALSE}
library(lubridate)
library(dplyr)
library(kableExtra)
library(formattable)

# setting
# set start time for the course (assumes same start time everyday)
time_begin <- as.POSIXct("09:00:00",format="%T")
```

```{r,eval=TRUE,include=FALSE}
s <- read.csv("schedule.csv",sep=";",stringsAsFactors=F,header=T,na.strings="",colClasses="character")

s <- s %>%
  # missing values in date and room are filled
  tidyr::fill(date,room,link_room,.direction="down") %>%
  # convert date characters to date objects
  dplyr::mutate(date=format(as_datetime(date,format="%d/%m/%Y"),"%d-%b-%Y")) %>%
  dplyr::mutate(day=format(as_datetime(date,format="%d-%b-%Y"),"%a")) %>%
  dplyr::group_by(date) %>%
  dplyr::mutate(end=time_begin+minutes(cumsum(dur)),start=c(time_begin,end[-n()])) %>%
  dplyr::mutate(start=format(start,"%H:%M"),end=format(end,"%H:%M")) %>%
  as.data.frame() %>%
  dplyr::mutate(room=ifelse(is.na(link_room),room,paste0("<a href='",link_room,"'>",room,"</a>"))) %>%
  # create day label (with date, day, room)
  dplyr::mutate(date=paste0("<p style='font-size:110%;padding-top:10px;padding-bottom:5px;margin-top:10px;'><span style='font-weight:bold;'><span style='color:#0093BD;margin-left:5px;margin-right:5px;'>",fontawesome::fa("calendar",fill=color_primary),"</span>",date,"</span>",
                     "<span style='color:white;border-radius:4px;padding-right:4px;padding-left:4px;background:#0093BD;margin-left:9px;margin-right:7px;'>",day,"</span>",
                     "<span style='color:#0093BD;margin-left:5px;margin-right:5px;'>",fontawesome::fa("map-marker-alt",fill=color_primary),"</span>",room,"</p>")) %>%
  # style topic
  dplyr::mutate(topic=ifelse(tolower(topic)=="coffee break" | tolower(topic)=="lunch",paste0("<span style='color:white;font-weight:bold;border-radius:4px;padding:4px;background:#BDBDBD;'>",topic,"</span>"),paste0("<span style='margin-right:5px;'>",topic,"</span>"))) %>%
  # add links to topic
  dplyr::mutate(topic=ifelse(is.na(link_slide),topic,paste0("<span>",topic,"<a href='",link_slide,"'><span style='margin-left:4px;margin-right:4px;'>",fontawesome::fa("file-powerpoint",fill=color_primary),"</span></a></span>"))) %>%
  mutate(topic=ifelse(is.na(link_lab),topic,paste0("<span>",topic,"<a href='",link_lab,"'><span style='margin-left:4px;margin-right:4px;'>",fontawesome::fa("file-alt",fill=color_primary),"</span></a></span>"))) %>%
  dplyr::mutate(time=paste0(start," - ",end)) %>%
  dplyr::mutate(teacher=ifelse(is.na(teacher),"",teacher)) %>%
  dplyr::mutate(assistant=ifelse(is.na(assistant),"",assistant)) %>%
  dplyr::select(date,room,time,topic,teacher,assistant)

if(all(is.na(s$assistant)) | all(s$assistant=="")) s$assistant <- NULL
```

<div class="table-schedule">

```{r,eval=TRUE,echo=FALSE,block.title=NULL}
# grouping vector
r <- rle(s$date)$length
names(r) <- rle(s$date)$values
cnames <- c("Time","Topic","Teacher","Assistant")

s %>%
  dplyr::select(-date,-room) %>%
  setNames(cnames[1:ncol(.)]) %>%
  kable(format="html",escape=FALSE) %>%
  kable_styling(bootstrap_options=c("hover","condensed"),position="center",full_width=TRUE) %>%
  column_spec(1,width="7em") %>%
  column_spec(2,width="16em") %>%
  pack_rows(index=r,label_row_css="background:white;",escape=F,indent=FALSE)
```

<!-- marker legend -->
<p class="small" style="margin-top:20px;">
<span>
<span style="margin-left:6px; margin-right:6px;">
<span style="margin-left:4px; margin-right:4px;">`r fontawesome::fa("calendar",fill=color_primary)`</span> Date
</span>
<span style="margin-left:6px; margin-right:6px;">
<span style="margin-left:4px; margin-right:4px;">`r fontawesome::fa("map-marker-alt",fill=color_primary)`</span> Room number
</span>
<span style="margin-left:6px; margin-right:6px;">
<span style="margin-left:4px; margin-right:4px;">`r fontawesome::fa("file-powerpoint",fill=color_primary)`</span> Slides
</span>
<span style="margin-left:6px; margin-right:6px;">
<span style="margin-left:4px; margin-right:4px;">`r fontawesome::fa("file-alt",fill=color_primary)`</span> Lab exercises
</span>
</span>
</p>

<p class="small" style="margin-top:20px;">
<span style="margin-left:4px; margin-right:4px;">`r fontawesome::fa("users",fill=color_primary)`</span> MD: Martin Dalhö, JD: Jacques Dainat, PE: Payam Emami, AS: Anders Sjölander, AJ: Anna Johansson, BS: Bengt Sennblad, NR: Nima Rafati, OP: Olga Petterson.
</p>
</div>
