---
title: "Practical Info"
subtitle: "Venue, address, travel and contact information"
date: ""
toc: false
number-sections: false
sidebar: false
format: html
---

```{r,eval=TRUE,include=FALSE}
library(yaml)
library(here)
library(leaflet)

#' Get location data from YAML into a list and data.frame
#' @param y YAML data
#' @param loc Name of location
#' @return List with description and data.frame with map data
#'
get_info <- function(y, loc) {
  loc <- tolower(loc)

  if (loc %in% names(y)) {
    info <- y[[loc]]
    if (!is.null(info$description)) {
      d <- info$description
    } else {
      d <- NULL
    }
    if (!is.null(info$map)) {
      m <- do.call(rbind, lapply(info$map, function(x){
        required_fields <- c("label", "marker_icon", "marker_color", "lat", "lon", "popup")
        missing_fields <- setdiff(required_fields, names(x))
        
        if (length(missing_fields) > 0) {
          stop(paste("Missing fields:", paste(missing_fields, collapse = ", ")))
        }
        as.data.frame(x)
      }))
    } else {
      m <- NULL
    }
    return(list(description = d, map = m))
  } else {
    message(paste0("Location ", loc, " not found in the data"))
    return(NULL)
  }
}

# set location in _quarto.yml
loc <- sapply(unlist(strsplit(yaml::read_yaml(here("_quarto.yml"))$location, ";|,")), trimws)
l <- yaml::yaml.load_file(here("info.yml"))
```

## Location

```{r,echo=FALSE,eval=TRUE,results="asis"}
cat("\n::: {.panel-tabset}\n")

for (i in seq_along(loc)) {
  p <- loc[i]
  cat(paste0("\n### ", tools::toTitleCase(p), "\n"))

  idata <- get_info(l, p)
  if (!is.null(idata)) {
    if (!is.null(idata$map)) {
      dfr1 <- idata$map
      dfr1$popup <- gsub(";|; ", "<br>", dfr1$popup)
      dfr2 <- dfr1[complete.cases(dfr1$lat, dfr1$lon), ]
      
      cat(knitr::knit_child(input = "assets/_child-info.qmd", envir = environment(), quiet = TRUE))
      cat("\n")
    }

    if (!is.null(idata$description)) {
      cat("\n", idata$description)
    }
  } else {
    cat("No additional information is provided.")
  }
  cat("\n")
}

cat(":::\n")
```

## Contact

This workshop is run by the **National Bioinformatics Infrastructure Sweden (NBIS)** in collaboration with **National Genomics Infrastructure (NGI)**. Both, NGI and NBIS are platforms at SciLifeLab.

If you would like to get in touch with us regarding this workshop, please contact us at **edu.intro-ngs [at] nbis.se**.

***

![](assets/logos/nbis-scilifelab-ngi.svg){height="26px"}
