---
title: "Practical Info"
subtitle: "Venue, address, travel and contact information"
date: ""
toc: false
number-sections: false
sidebar: false
format: html
---

```{r,eval=TRUE,include=FALSE}
library(yaml)
library(here)
library(leaflet)

# set location in _quarto.yml
loc <- unlist(strsplit(yaml::read_yaml(here("_quarto.yml"))$location,";|,"))
dfr <- read.delim("info.csv",sep=";")
```

## Location

```{r,echo=FALSE,results="asis",eval=TRUE}
cat("::: {.panel-tabset}", "\n")

p <- "uppsala"
cat(paste0("\n### ",tools::toTitleCase(p),"\n"))
dfr1 <- dfr[dfr$location %in% p,,drop=FALSE]
dfr1$address <- gsub(",","<br>",dfr1$address)
icons <- awesomeIcons(icon=dfr1$marker_icon, markerColor=dfr1$marker_colour, iconColor="#FFFFFF", library="fa")
htmltools::tagList(leaflet(height=450, elementId=paste0("leaflet-",p),width="auto") %>%
  addTiles(urlTemplate='http://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png') %>%
  addAwesomeMarkers(lat=dfr1$lat, lng=dfr1$lon, label=dfr1$label, popup=dfr1$address, icon=icons) %>%
  setView(lat=dfr1$lat[1], lng=dfr1$lon[1], zoom=13))
cat(paste0("\n",dfr1$address[1],"\n"))
cat(paste0("\n",dfr1$description[1],"\n"))

# p <- "umea"
# cat(paste0("\n### ",tools::toTitleCase(p),"\n"))
# dfr1 <- dfr[dfr$location %in% p,,drop=FALSE]
# dfr1$address <- gsub(",","<br>",dfr1$address)
# icons <- awesomeIcons(icon=dfr1$marker_icon, markerColor=dfr1$marker_colour, iconColor="#FFFFFF", library="fa")
# htmltools::tagList(leaflet(height=450, elementId=paste0("leaflet-",p),width="auto") %>%
#   addTiles(urlTemplate='http://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png') %>%
#   addAwesomeMarkers(lat=dfr1$lat, lng=dfr1$lon, label=dfr1$label, popup=dfr1$address, icon=icons) %>%
#   setView(lat=dfr1$lat[1], lng=dfr1$lon[1], zoom=14))
# cat(paste0("\n",dfr1$address[1],"\n"))
# cat(paste0("\n",dfr1$description[1],"\n"))

# p <- "lund"
# cat(paste0("\n### ",tools::toTitleCase(p),"\n"))
# dfr1 <- dfr[dfr$location %in% p,,drop=FALSE]
# dfr1$address <- gsub(",","<br>",dfr1$address)
# icons <- awesomeIcons(icon=dfr1$marker_icon, markerColor=dfr1$marker_colour, iconColor="#FFFFFF", library="fa")
# htmltools::tagList(leaflet(height=450, elementId=paste0("leaflet-",p),width="auto") %>%
#   addTiles(urlTemplate='http://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png') %>%
#   addAwesomeMarkers(lat=dfr1$lat, lng=dfr1$lon, label=dfr1$label, popup=dfr1$address, icon=icons) %>%
#   setView(lat=dfr1$lat[1], lng=dfr1$lon[1], zoom=14))
# cat(paste0("\n",dfr1$address[1],"\n"))
# cat(paste0("\n",dfr1$description[1],"\n"))

cat(":::", "\n")
```

```{r,echo=FALSE,eval=FALSE,results="asis"}
if (!"online" %in% loc) {
    cat("\n::: {.panel-tabset}\n")

    for (i in seq_along(loc)) {
        p <- loc[i]
        cat(paste0("\n### ", tools::toTitleCase(p), "\n"))
        dfr1 <- dfr[dfr$location %in% p, , drop = FALSE]
        dfr1$address <- gsub(",", "<br>", dfr1$address)
        icons <- awesomeIcons(icon = dfr1$marker_icon, markerColor = dfr1$marker_colour, iconColor = "#FFFFFF", library = "fa")

        x <- htmltools::tagList(leaflet(height = 450, width = "auto", elementId=paste0("leaflet-",p)) %>%
            addTiles(urlTemplate = "http://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png") %>%
            addAwesomeMarkers(lat = dfr1$lat, lng = dfr1$lon, label = dfr1$label, popup = dfr1$address, icon = icons) %>%
            setView(lat = dfr1$lat[1], lng = dfr1$lon[1], zoom = 13))
        
        cat(as.character(x))
        cat("\n")
        cat("\n", dfr1$address[1], "\n")
        cat("\n", dfr1$description[1], "\n")
    }

    cat(":::\n")
}else{
  cat("This workshop is online only. The online meeting details will be be sent to participants by email.\n")
}
```

## Contact

This workshop is run by the **National Bioinformatics Infrastructure Sweden (NBIS)** in collaboration with **National Genomics Infrastructure (NGI)**. Both, NGI and NBIS are platforms at SciLifeLab.

If you would like to get in touch with us regarding this workshop, please contact us at **edu.intro-ngs [at] nbis.se**.

***

![](assets/logos/nbis-scilifelab-ngi.svg){height="26px"}
