---
title:  "Status & Troubleshooting"
description: "Utility commands/scripts, troubleshooting, tips and suggestions."
date: ""
sidebar: false
format: html
---

```{r}
#| echo: false
library(yaml)
library(here)
y <- yaml::read_yaml(here("_quarto.yml"))
upid <- y$uppmax_project
upres1 <- y$uppmax_res_1
upres2 <- y$uppmax_res_2
upres3 <- y$uppmax_res_3
upres4 <- y$uppmax_res_4
upres5 <- y$uppmax_res_5
rm(y)
```

## Status

Useful commands for monitoring status and user activity on UPPMAX.

### List jobs

List all jobs running an a project

```{r}
#| echo: false
#| class-output: bash
cat(paste0("squeue -A ",upid," | sort -k 4"))
```

Sample output

```
  23186770      core _interac analopez  R    2:15:00      1 r55
  23175660      core   (null)   anitav  R    6:52:28      1 r480
  23188713      core   (null)    annha  R    1:42:59      1 r480
  23185442      core   (null)    annha  R    2:47:11      1 r480
  23185433      core   (null) inene424  R    2:47:42      1 r479
  23185391      core   (null) inene424  R    2:50:04      1 r479
  23185540      core   (null)  shengyz  R    2:46:20      1 r480
     JOBID PARTITION     NAME     USER ST       TIME  NODES NODELIST(REASON)
  23185119      core   (null)    wingf  R    2:53:40      1 r479
```

### Jobs per user

Sorted list of jobs per user

```{r}
#| echo: false
#| class-output: bash
cat(paste0("squeue -A ",upid," | tr -s ' ' | cut -d' ' -f5 | sort | uniq -c | sort -k1"))
```

Sample output

```
      1 tami
      1 tommal
      1 USER
      1 valeriia
      1 vioww
      1 vishnupk
      1 ylvafr
      2 mehran96
      2 miika
      3 mariasve

```

### Core usage

Total number of cores used on a project

```{r}
#| echo: false
#| class-output: bash
cat(paste0("squeue -A ",upid," -o %C | awk '{total += $0} END{print total}'"))
```

### Space usage

Amount of storage space used per project

```{r}
#| echo: false
#| class-output: bash
cat(paste0("uquota | grep ",upid))
```

Sample output

```
Your project     Your File Area         Unit       Usage  Quota Limit  Over Quota
---------------  ---------------------  --------  ------  -----------  ----------
snic2022-22-123  /proj/snic2022-22-123  GBytes       4.6          128
```

### Reservation use

List users by reservation ID

```{r}
#| echo: false
#| class-output: bash
cat(paste0("squeue -R ",upres1," | sort -k 4\n"))
cat(paste0("squeue -R ",upres2," | sort -k 4\n"))
cat(paste0("squeue -R ",upres3," | sort -k 4\n"))
cat(paste0("squeue -R ",upres4," | sort -k 4\n"))
cat(paste0("squeue -R ",upres5," | sort -k 4\n"))
```

Sample output

```
23188713      core   (null)    annha  R    2:01:15      1 r480
23185442      core   (null)    annha  R    3:05:27      1 r480
23187840      core   (null)   evangc  R    2:11:02      1 r479
23186811      core   (null)  shengyz  R    2:34:45      1 r479
23185540      core   (null)  shengyz  R    3:04:36      1 r480
   JOBID PARTITION     NAME     USER ST       TIME  NODES NODELIST(REASON)
23185119      core   (null)    wingf  R    3:11:56      1 r479

```

### User activity

List last activity in a directory for all users

```{r}
#| echo: false
#| class-output: bash
cat(paste0("bash /sw/courses/utils/list_modification_times.sh /proj/",upid,"/nobackup/"))
```

Sample output

```
hkyle           2021-11-22 13:59:59     (/proj/snic2021-22-644/nobackup/hkyle/uppmax_tutorial/job_template)
malinh          2021-11-22 14:12:10     (/proj/snic2021-22-644/nobackup/malinh/slurm-23182638.out)
aliraz          2021-11-22 14:23:16     (/proj/snic2021-22-644/nobackup/aliraz/slurm-23184022.out)
kristaps        2021-11-22 14:25:08     (/proj/snic2021-22-644/nobackup/kristaps/uppmax_tutorial/jobData.sam)
anapin          2021-11-22 14:56:14     (/proj/snic2021-22-644/nobackup/anapin/uppmax_tutorial/uppmax_tutorial/jobData.sam)
analopez        2021-11-22 14:56:16     (/proj/snic2021-22-644/nobackup/analopez/uppmax_tutorial/jobData.sam)
private         Not available     ()
```

### List reservations for a project

```{r}
#| echo: false
#| class-output: bash
cat(paste0("scontrol show res | grep ",upid))
```

Sample output

```
ReservationName=snic2021-22-644_wed StartTime=2021-11-24T12:00:00 EndTime=2021-11-24T18:00:00 Duration=06:00:00
   Users=(null) Accounts=snic2021-22-644 Licenses=(null) State=INACTIVE BurstBuffer=(null) Watts=n/a
ReservationName=snic2021-22-644_thu StartTime=2021-11-25T08:30:00 EndTime=2021-11-25T17:30:00 Duration=09:00:00
   Users=(null) Accounts=snic2021-22-644 Licenses=(null) State=INACTIVE BurstBuffer=(null) Watts=n/a
ReservationName=snic2021-22-644_fri StartTime=2021-11-26T08:30:00 EndTime=2021-11-26T13:30:00 Duration=05:00:00
   Users=(null) Accounts=snic2021-22-644 Licenses=(null) State=INACTIVE BurstBuffer=(null) Watts=n/a
```

### List project members

```{r}
#| echo: false
#| class-output: bash
cat(paste0("projmembers ",upid))
```

Sample output

```
snic2022-22-123 abusiddi   Abu Bakar Siddique
snic2022-22-123 afifac     Afifa Enam Chowdhury
snic2022-22-123 asve0014   Ashish Verma
snic2022-22-123 berka      Berkay Paylar
snic2022-22-123 btleren    Betül Eren Keskin
```

### User identity

Fetch user information from username

```{bash}
#| eval: false
finger username
```

## Troubleshooting

### X-forwarding

#### Setup

**Mac users**

- Install Xquartz and restart computer
- Open the Xquartz terminal and connect to Uppmax 
- In my recent macos (10.xxx) , x-forwarding only works if I explicitly run the Xquartz terminal

`ssh -XY username@rackham.uppmax.uu.se`

Also use this when logging in to the compute node!

`ssh -XY username@node`

**Windows users**

In MobaXTerm, go to settings and make sure that X-forwarding is checked.

#### Testing X-forwarding

Type `xeyes` in the terminal.

### Open .html documents on Rackham

First ensure X-forwarding works, then run `firefox --no-remote filename.html`

### SCP fails with *

Sometimes students have problems to download files with SCP when there is a `*` in the end of the line. For example; `scp user@rackham.uppmax.uu.se:/proj/bla/HG00097.bam* .`. It needs to be changed to `scp user@rackham.uppmax.uu.se:/proj/bla/HG00097.bam\* .`.

### Thinlinc

#### Login

When logging in through an installed client, **username/password** seems to work. When logging in through the browser, **username/password+2FA** may be required.

#### Minimize the ThinLinc window

On a  Mac, press <kbd>Fn + F8</kbd>, then select: Minimize window. 

#### Gedit opening issues

When opening it from the terminal (`gedit &`), it was not able to connect to a display to show the graphics. The `DISPLAY` variable was empty. This was when using ThinLinc, so it should have worked. The issue was solved by opening gedit from the menu. Gedit could be labelled Text Editor. 

### Set persistent home directory in MobaXTerm

This is to specify where the **home directory** in MobaXTerm is located in the computer’s file system. In **Settings > Configuration**, set the persistent home directory to a suitable folder. Restart MobaXTerm. 

### Typing `$` on a Swedish keyboard

Press <kbd>AltGr + 4</kbd>.

### Black background for XQuartz windows on M1 Mac

As documented [here](https://github.com/XQuartz/XQuartz/issues/31), run:

```{bash}
#| eval: false
defaults write org.xquartz.X11 enable_render_extension 0
```

***
