---
title: "Uppmax IO"
subtitle: "Downloading and uploading files from Uppmax"
date: ""
format: html
---

```{r,eval=TRUE,include=FALSE}
library(yaml)
library(here)
id_project <- yaml::read_yaml(here("_quarto.yml"))$id_project
path_workspace <- yaml::read_yaml(here("_quarto.yml"))$path_workspace
url_cluster <- yaml::read_yaml(here("_quarto.yml"))$url_cluster
```

## Local workspace

Start by creating a folder on your laptop where the files that you will download should end up. You need to have write permission in this folder. This folder will be referred to as your *local workspace* throughout these instructions.

Open a terminal window on your laptop and move into your local workspace.

## Download a file from UPPMAX

Lets assume that you have a file "results.txt" in the following folder on UPPMAX:

```{r}
#| echo: false
#| class-output: bash
cat(paste0(path_workspace,'/somefolder/'))
```

::: {.callout-note}
`username` and `somefolder` should be replaced with your real username and a real folder name.
:::

To download the file to your local workspace type:

```{r}
#| output: asis
#| echo: false
out <- paste0("scp username@", url_cluster, ":", path_workspace, "/somefolder/results.txt .")
cat(
  "```{.bash filename='Local Terminal'}",
  out,
  "```",
  sep = "\n"
)
```

Note that the last `.` means that the file will keep the original name.

## Upload a file to UPPMAX

Now lets imagine that you have developed a script on your laptop and want to upload it to UPPMAX. The script is stored in your local workspace and is called "script.sh". Type this in your local workspace to upload the file to UPPMAX:

::: {.callout-note}
`username` and `somefolder` should be replaced with your real username and a real folder name.
:::

```{r}
#| output: asis
#| echo: false
out <- paste0("scp script.sh username@", url_cluster, ":", path_workspace, "/somefolder/.")
cat(
  "```{.bash filename='Local Terminal'}",
  out,
  "```",
  sep = "\n"
)
```
